pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-08-06 bmc_article.tex
Updates from Overleaf

Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition ({\color{red}Table S\ref{tab:table1}}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see {\color{red}Table S\ref{tab:table1}} and Additional File 1: Fig S1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{{normalization.pdf}}
    \caption{\textbf{Evaluation of normalization procedures. A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster. Grey squares indicate that the true number of clusters could not be reached with the corresponding method, dataset and parameter set.} `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling. Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping. Other evaluation metric heatmaps (e.g. panel \textbf{C} here) map colors to the signed square-root of the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median. Using this mapping, \textit{differences} in color have the same meaning across datasets, and the color-scale is not primarily capturing outliers or baseline differences between datasets. The numbers printed in the cells represent the raw (i.e. unscaled) metric values, and are printed in white or black depending on whether they are above or below the median.
    \label{fig:norm}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.55\textwidth, keepaspectratio]{{anova.pdf}}
    \caption{\color{red}\textbf{Variance in clustering metrics explained by the parameters.} Proportion of variance explained by the parameters of the main analysis steps. The proportions are calculated from an ANOVA using the model;
    \texttt{metric $\sim$ dataset*resolution+doubletmethod+filt+norm+clustmethod+dims}
    and substracting the variance attributable to differences due to datasets, resolutions and their interactions. }
    \label{fig:overall}
\end{figure}

@@

\begin{table}[h!]
\caption{{\color{red} \textbf{Table S1:} Overview of the benchmark datasets.} \# features' indicates the number of features detected in at least 10 cells.}
\label{tab:table1}
\begin{tabular}{rlrrl}
  \hline
dataset & \# features & \# cells & protocol & description \\ 
  \hline
Koh & 33922 & 531 & SMARTer & 9 FACS purified differentiation stages \\ 
  Kumar & 41930 & 246 & SMARTer & Mouse ESC cultured in 3 different conditions \\ 
  Zhengmix4eq & 10434 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix4uneq & 11369 & 6498 & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix8eq & 10600 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
  mixology10x3cl & 16208 & 902 & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
  mixology10x5cl & 11786 & 3918 & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
  simMix1 & 3696 & 2500 & 10x-based & Simulation of 10 human cell subpopulations \\
  simMix2 & 8893 & 3000 & 10x-based & Simulation of 9 mouse cell subpopulations \\
   \hline
\end{tabular}
\end{table}

>>>>

Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition ({\color{red}Table \ref{tab:table1}}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see {\color{red}Table \ref{tab:table1}} and Additional File 1: Fig S1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{{normalization.pdf}}
    \caption{\textbf{Evaluation of normalization procedures. A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster. Grey squares indicate that the true number of clusters could not be reached with the corresponding method, dataset and parameter set. `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling. Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping. Other evaluation metric heatmaps (e.g. panel \textbf{C} here) map colors to the signed square-root of the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median. Using this mapping, \textit{differences} in color have the same meaning across datasets, and the color-scale is not primarily capturing outliers or baseline differences between datasets. The numbers printed in the cells represent the raw (i.e. unscaled) metric values, and are printed in white or black depending on whether they are above or below the median.}
    \label{fig:norm}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.55\textwidth, keepaspectratio]{{anova.pdf}}
    \caption{\color{red}\textbf{Variance in clustering metrics explained by the parameters.} Proportion of variance explained by the parameters of the main analysis steps. The proportions are calculated from an ANOVA using the model \hspace{\textwidth}
    \texttt{metric $\sim$  dataset*resolution+doubletmethod+filt+norm+clustmethod+dims} \hspace{\textwidth}
    and substracting the variance attributable to differences due to datasets, resolutions and their interactions. }
    \label{fig:overall}
\end{figure}

@@

\begin{table}[h!]
\caption{Overview of the benchmark datasets. \# features' indicates the number of features detected in at least 10 cells.}
\label{tab:table1}
\begin{tabular}{rlrrl}
  \hline
dataset & \# features & \# cells & protocol & description \\ 
  \hline
Koh & 33922 & 531 & SMARTer & 9 FACS purified differentiation stages \\ 
  Kumar & 41930 & 246 & SMARTer & Mouse ESC cultured in 3 different conditions \\ 
  Zhengmix4eq & 10434 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix4uneq & 11369 & 6498 & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix8eq & 10600 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
  mixology10x3cl & 16208 & 902 & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
  mixology10x5cl & 11786 & 3918 & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
  simMix1 & 3696 & 2500 & 10x-based & Simulation of 10 human cell subpopulations \\
  simMix2 & 8893 & 3000 & 10x-based & Simulation of 9 mouse cell subpopulations \\
   \hline
\end{tabular}
\end{table}
